---
title: "USP 634 Lab 3- Some R Basics"
author: "Jamaal Green and Liming Wang"
output: 
  html_document:
    toc: true
    self_contained: true
    number_sections: true
---


# Run Simulations with R

Simulate random draws from a population:

> sample(population, size (the number of items to choose), replace = FALSE)

Repeat the simulation with:

> replicate(M, function)

## Roll dices

We can combine **sample()** and **replicate()** to simulate the rolling of dice. 

```{r, warning=FALSE, message=FALSE, eval=FALSE}
#Simulate rolling a dice: 
sample(1:6, 1)

#Check whether a single roll gets a 4
sample(1:6, 1) == 4

#Roll two die and check if we get a 4 and 6

sample(1:6, 1) == 4 & sample(1:6, 1) == 4
```

```{r, warning=FALSE, message=FALSE}
#Now we will repeatedly roll two dices for 10000 times, check wether we get 4 and 6,
#save it into an object called rep10k

rep10k <- replicate(10000, sample(1:6, 1) == 4 & sample(1:6, 1) == 6)

#Calculate probability by summing up our
#TRUE responses and then dividing by the 
#number of observations (10000)

sum(rep10k, TRUE)/10000
```

## Verify Central Limit Theorem with Simulations

```{r}
library(dplyr)
library(ggplot2)
## Assume a uniform distribution of 100 numbers from 1 to 100 
population <- 1:100

## randomly sample 10 with replacement and calculate their sample mean
sample(population, 10, replace=TRUE) %>% mean

## repeatedly do a sample of 10 for 1000 times
clt_rep1k <- replicate(1000, sample(population, 10, replace=TRUE) %>% mean)

## Let's plot the sample means for the 1000 samples
qplot(clt_rep1k)

## Verify by plotting the normal curve of the sampling distribution on top of the histogram
ggplot(data.frame(x=clt_rep1k), aes(x)) + 
  geom_histogram(aes(y=..density..)) + 
  stat_function(fun=dnorm, args=list(mean=50, sd=sd(1:100)/sqrt(10)))
```

# Confidence Intervals


Data File
    - [californiatod.csv](californiatod.csv)

    
```{r}
#install.packages(readr)
library(readr)
library(dplyr)
californiatod <- read_csv("californiatod.csv")
```    

## Manual Calculations

First, compute the 95% **confidence intervals** for density by hand and looking up the z value from the `t` table, using R as a calculator. 

```{r, warning=FALSE, message=FALSE}

xbar <- mean(californiatod$density)
s <- sd(californiatod$density)
n <- nrow(californiatod)
alpha <- 1 - 0.95
tstar <- qt(1 - alpha/2, df = n-1)
SE <- s/sqrt(n)
MOE <- tstar*SE
xbar + c(-1,1) * MOE
```

## CI from t.test

You can get the confidence interval when you run a single sample t-test in R (to be covered next week) using the command 

```
t.test(californiatod$density, alternative='two.sided', mu=0, conf.level=.95)
```

Verify whether your results are the same using these 2 difference methods.